% sections/abstract.tex — Abstract
\begin{abstract}
Synthetic data generation for enterprise accounting and financial auditing
remains a critical bottleneck for machine-learning research, fraud-detection
system development, and regulatory compliance testing.  Existing approaches
either produce statistically na\"ive tabular data that fails basic accounting
invariants or require access to sensitive production datasets that cannot be
shared.  We present \system{}, an open-source framework that generates
\emph{end-to-end coherent} synthetic enterprise data spanning journal entries,
document flows, subledgers, financial statements, intercompany eliminations,
and process-mining event logs---all from a single declarative YAML
configuration.

\system{} makes five main contributions.
\textbf{(i)}~A \emph{statistically grounded generation engine} built on
log-normal mixture models, five copula families (Gaussian, Clayton, Gumbel,
Frank, Student-$t$), and Benford's-law--compliant amount sampling that
preserves cross-field correlations and marginal properties simultaneously.
\textbf{(ii)}~A \emph{multi-stage anomaly injection framework} with 33
anomaly types organized across five categories (fraud, error, process,
statistical, relational), including progressive fraud schemes with
state-machine advancement, configurable difficulty, severity, and confidence
scoring.
\textbf{(iii)}~A \emph{privacy-preserving fingerprint module} that extracts
distributional summaries from real datasets under differential privacy
($\varepsilon$-DP) and $k$-anonymity, enabling synthetic data that mirrors
production characteristics without exposing individual records.
\textbf{(iv)}~A \emph{comprehensive evaluation suite} comprising 15
domain-specific coherence validators, Benford and distribution-fit tests,
temporal-pattern analysis, drift detection, and ML-readiness metrics including
graph-neural-network feature assessment.
\textbf{(v)}~A \emph{counterfactual simulation engine} with causal DAG
propagation for what-if analysis, enabling controlled interventions on
financial process dependencies and assessment of downstream impacts.

Implemented in Rust across 15 crates totaling over 100\,k lines of code,
\system{} achieves throughput exceeding $2 \times 10^5$ journal entries per
second on a single core with deterministic reproducibility via
ChaCha8-seeded RNG.  The framework supports five accounting frameworks
(US~GAAP, IFRS, French~GAAP, German~GAAP, and dual-reporting) and
multi-period stateful generation with checkpoint support for long-horizon
simulations.
We demonstrate that synthetic datasets generated by \system{} satisfy
the accounting equation, preserve Benford's-law compliance (MAD $< 0.006$),
maintain document-chain referential integrity, and support downstream
fraud-detection models that achieve within 3\% of the F1 scores obtained on
real data.

\medskip
\noindent\textbf{Keywords:} synthetic data generation, enterprise accounting,
fraud detection, Benford's law, copula models, differential privacy,
process mining, graph neural networks, audit analytics, counterfactual
analysis, causal inference
\end{abstract}
