% sections/architecture.tex — System Architecture
\section{System Architecture}
\label{sec:architecture}

\system{} is implemented as a Rust workspace comprising 16 crates
organized into five architectural layers (\Cref{fig:architecture}).  The
layered design enforces separation of concerns and maps directly onto
the three-layer knowledge model (\Cref{def:three_layers}): domain models
and statistical primitives in the foundation layer encode Layer~2
(statistical) knowledge, generators in the process layer produce Layer~1
(structural) knowledge, and standards and evaluation modules encode
Layer~3 (normative) knowledge.

% ─────────────────────────────────────────────────────────────────────
\subsection{Layered Crate Organization}
\label{sec:arch_layers}

\begin{figure}[t]
\centering
\fbox{\parbox{0.92\textwidth}{%
\small
\begin{verbatim}
Layer 5  UI / Interfaces
         datasynth-cli | datasynth-server | datasynth-ui

Layer 4  Analytics & Export
         datasynth-eval | datasynth-graph | datasynth-graph-export
         datasynth-fingerprint

Layer 3  Domain Generators
         datasynth-generators | datasynth-banking | datasynth-ocpm
         datasynth-standards

Layer 2  Orchestration & Output
         datasynth-runtime | datasynth-output | datasynth-config

Layer 1  Foundation
         datasynth-core (models, distributions, resource guards)
         datasynth-test-utils
\end{verbatim}
}}
\caption{Layered crate architecture of \system{} (16 crates).
  Dependency arrows point downward; each layer depends only on layers
  below it.}
\label{fig:architecture}
\end{figure}

\paragraph{Layer 1: Foundation.}
\texttt{datasynth-core} defines domain model types spanning
accounting (journal entries, ACDOCA), master data (vendors, customers,
materials, fixed assets, employees), document flows (P2P, O2C),
sourcing (S2C), financial reporting, HR/payroll, manufacturing, sales,
bank reconciliation, intercompany, subledgers, FX/close,
tax accounting, treasury management, ESG/sustainability, project
accounting, audit (ISA~600 group audit), audit documentation and
methodology, business combinations, pensions, and relationship/graph
types.  Statistical distribution samplers include Benford, copula
(5 families), mixture, Pareto, Weibull, Beta, zero-inflated,
conditional, drift, and temporal models.  Resource-management primitives
(memory guard, CPU monitor, disk guard, degradation controller) ensure
stable operation under load.  All financial values use
\texttt{rust\_decimal::Decimal} for exact arithmetic, eliminating
IEEE~754 rounding errors.

\paragraph{Layer 2: Orchestration.}
\texttt{datasynth-runtime} houses the \texttt{GenerationOrchestrator}
that coordinates the full generation workflow.
\texttt{datasynth-config} provides schema validation, industry presets
(manufacturing, retail, financial services, healthcare, technology), and
complexity-level mappings.
\texttt{datasynth-output} implements streaming sinks for CSV, JSON,
Parquet, and Apache Arrow formats.

\paragraph{Layer 3: Domain Generators.}
\texttt{datasynth-generators} contains specialized generators
organized by business process: P2P document flows, O2C chains, S2C
sourcing, HR/payroll (including benefits enrollment), manufacturing
(BOM, quality, cycle counts, inventory movements), tax accounting
(multi-jurisdiction, VAT/GST, deferred tax, transfer pricing),
treasury (cash positioning, forecasting, hedging, debt covenants),
ESG/sustainability (GHG emissions, energy, water, waste, diversity,
safety, governance), project accounting (WBS, earned value, change
orders, milestones), period-close, intercompany, and anomaly injection.
\texttt{datasynth-banking} adds KYC/AML-specific generation with five
money laundering typologies.
\texttt{datasynth-ocpm} produces OCEL~2.0 compliant event logs.
\texttt{datasynth-standards} implements five accounting frameworks
(US~GAAP, IFRS, French GAAP/PCG, German GAAP/HGB, and dual-reporting)
together with ISA, SOX, and PCAOB audit standard models.

\paragraph{Layer 4: Analytics.}
\texttt{datasynth-eval} provides the evaluation framework (15+
coherence validators, statistical tests, ML-readiness metrics,
auto-tuner).
\texttt{datasynth-graph} and \texttt{datasynth-graph-export} construct
and export heterogeneous reference knowledge graphs in PyTorch Geometric,
Neo4j, and DGL formats with 78+ entity types and 39+ relationship edge
types.
\texttt{datasynth-fingerprint} implements privacy-preserving extraction
and synthesis.

\paragraph{Layer 5: Interfaces.}
A CLI binary (\texttt{datasynth-data}) supports \texttt{generate},
\texttt{validate}, \texttt{init}, \texttt{info}, and \texttt{fingerprint}
subcommands.  A REST/gRPC/WebSocket server enables streaming generation
with API-key authentication and rate limiting.  A Tauri + SvelteKit
desktop UI provides visual configuration and monitoring.

% ─────────────────────────────────────────────────────────────────────
\subsection{Knowledge Layer Mapping}
\label{sec:arch_knowledge_mapping}

The three knowledge layers (\Cref{def:three_layers}) map to the
architecture as follows:

\begin{center}
\small
\begin{tabular}{lll}
\toprule
\textbf{Knowledge Layer} & \textbf{Crates} & \textbf{Artifacts} \\
\midrule
Structural ($\mathcal{K}_S$) &
  generators, graph, ocpm &
  Entity types, edges, document chains \\
Statistical ($\mathcal{K}_\Sigma$) &
  core (distributions), eval &
  Amount distributions, correlations, temporal \\
Normative ($\mathcal{K}_N$) &
  standards, generators (controls) &
  GAAP/IFRS rules, ISA/SOX, COSO \\
\bottomrule
\end{tabular}
\end{center}

Every generated record carries provenance tracing it to one or more
knowledge layers, enabling downstream consumers to understand not just
\emph{what} was generated but \emph{why}---a property that is difficult to achieve comprehensively when
constructing knowledge from observed data alone.

% ─────────────────────────────────────────────────────────────────────
\subsection{Deterministic Reproducibility}
\label{sec:arch_determinism}

All randomness flows from a single \texttt{u64} seed through the ChaCha8
cryptographic PRNG~\citep{bernstein2008chacha}.  Each generator receives
a derived seed computed as $\texttt{base\_seed} + \texttt{offset}(g)$
where $\texttt{offset}(g)$ is a fixed constant per generator type $g$.
This guarantees:

\begin{enumerate}
  \item \textbf{Bitwise reproducibility:} identical seed $\Rightarrow$
    identical output across runs and platforms.
  \item \textbf{Generator independence:} modifying one generator's
    parameters does not perturb another's output stream.
  \item \textbf{Parallel safety:} each thread maintains a local RNG
    instance; no global state is shared.
\end{enumerate}

Deterministic reproducibility is essential for reference knowledge
graphs: the same seed always produces the same ground truth, enabling
controlled experiments and reproducible benchmarks.

% ─────────────────────────────────────────────────────────────────────
\subsection{Collision-Free UUID Generation}
\label{sec:arch_uuid}

Synthetic records require globally unique identifiers that are
deterministic yet collision-free across generator types.  \system{}
employs an FNV-1a hash-based UUID factory with a 16-byte structure:

\begin{equation}
\underbrace{\texttt{seed}[0{:}5]}_{\text{6 bytes}}
\;\|\;
\underbrace{\texttt{gen\_type}}_{\text{1 byte}}
\;\|\;
\underbrace{\texttt{0x4} \| \texttt{sub}}_{\text{1 byte}}
\;\|\;
\underbrace{\texttt{counter}}_{\text{8 bytes}}
\label{eq:uuid_structure}
\end{equation}

The \texttt{gen\_type} discriminator byte (39 values,
$\texttt{0x01}$--$\texttt{0x3E}$) partitions the UUID space by
generator, ensuring that journal-entry IDs can never collide with vendor
IDs or document-flow IDs regardless of counter values.

% ─────────────────────────────────────────────────────────────────────
\subsection{Resource Management and Graceful Degradation}
\label{sec:arch_resources}

Large-scale generation runs may stress system resources.  \system{}
implements a unified resource guard monitoring memory (via
\texttt{/proc/self/statm} on Linux, \texttt{ps} on macOS), disk space
(via \texttt{statvfs}), and CPU utilization with configurable thresholds.
When thresholds are breached, the system transitions through four
degradation levels:

\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Level} & \textbf{Memory} & \textbf{Batch Size}
  & \textbf{Features Disabled} \\
\midrule
Normal     & $<70\%$  & $100\%$ & None \\
Reduced    & $70{-}85\%$ & $50\%$  & Data quality injection \\
Minimal    & $85{-}95\%$ & $25\%$  & Anomaly injection \\
Emergency  & $>95\%$  & Flush   & Optional fields \\
\bottomrule
\end{tabular}
\end{center}

A 5\% hysteresis band prevents oscillation between levels.

% ─────────────────────────────────────────────────────────────────────
\subsection{Configuration System}
\label{sec:arch_config}

\system{} uses a hierarchical YAML configuration with 30+ sections
covering global settings, companies, chart of accounts, transactions,
document flows, distributions, temporal patterns, anomaly injection,
internal controls, intercompany, banking, accounting and audit
standards, vendor networks, customer segmentation, relationship
strength, cross-process links, tax, treasury, ESG, project accounting,
HR, manufacturing, sales, and more.

Five industry presets (manufacturing, retail, financial services,
healthcare, technology) provide complete default configurations.
Three complexity levels (small $\sim$100 accounts, medium $\sim$400,
large $\sim$2{,}500) control chart-of-accounts granularity and
transaction volume.

\paragraph{Multi-GAAP Framework.}
\system{} supports framework-aware generation across five accounting
standards: US~GAAP, IFRS, French GAAP (Plan Comptable
G\'{e}n\'{e}ral), German GAAP (HGB), and dual-reporting mode.  Each
framework determines GL account numbering conventions, classification
rules, and depreciation methods.  Seven country packs (FR, JP, CN, IN,
IT, ES, CA) bundle currency, tax regime, fiscal-year convention, and
banking configurations.  German jurisdictions support GoBD-compliant
audit export, and French jurisdictions produce FEC extracts.

% ─────────────────────────────────────────────────────────────────────
\subsection{Streaming Pipeline and Session Management}
\label{sec:arch_streaming}

\system{} supports phase-aware streaming output via the
\texttt{PhaseSink} trait.  Three sink implementations are provided:
file (JSONL), HTTP, and no-op (benchmarking).

Multi-period generation is managed by \texttt{GenerationSession}, which
persists state to \texttt{.dss} checkpoint files between runs.
\texttt{SessionState} tracks RNG seeds, account balance snapshots,
document-ID counters, and entity counts across periods, enabling
incremental generation and long-horizon simulations.

% ─────────────────────────────────────────────────────────────────────
\subsection{Performance}
\label{sec:arch_performance}

Key optimizations include \texttt{SmallVec<[T; 4]>} for line-item
collections, precomputed temporal CDF tables, binary search over
cumulative weights, \texttt{itoa}/\texttt{ryu} fast formatting, and
\texttt{zstd} output compression.  The \texttt{ParallelGenerator}
trait splits the seed space deterministically across threads,
achieving near-linear scaling up to 8 cores and exceeding
$2 \times 10^5$ journal entries per second on a modern x86-64 core.
