% sections/experiments.tex — Experiments
\section{Experiments}
\label{sec:experiments}

We evaluate \system{} along four dimensions: statistical fidelity,
structural coherence, downstream ML utility, and system performance.

% ─────────────────────────────────────────────────────────────────────
\subsection{Experimental Setup}
\label{sec:exp_setup}

\paragraph{Configurations.}
We generate datasets at three complexity levels using industry presets:

\begin{center}
\small
\begin{tabular}{llrrr}
\toprule
\textbf{Preset} & \textbf{Complexity} & \textbf{GL Accounts}
  & \textbf{Entities} & \textbf{Period} \\
\midrule
Retail          & Small  & $\sim$100  & 1  & 12 months \\
Manufacturing   & Medium & $\sim$400  & 3  & 24 months \\
Financial Svcs. & Large  & $\sim$2{,}500 & 10 & 36 months \\
\bottomrule
\end{tabular}
\end{center}

Each configuration is run with anomaly injection enabled at rates of
2\%, 5\%, and 10\%.  All experiments use seed $= 42$ for reproducibility.

\paragraph{Hardware.}
All benchmarks are run on a single machine with an AMD Ryzen 9 7950X
(16 cores / 32 threads), 64\,GB RAM, and NVMe SSD, running Linux 6.x.

% ─────────────────────────────────────────────────────────────────────
\subsection{Grounding in Real-World Accounting Data}
\label{sec:exp_realworld}

To validate that \system{} produces data consistent with real-world
enterprise accounting, we compare generated characteristics against
empirical distributions observed in production general ledger data.
\citet{ivertowski2024hardware} analyzed 155 real-world
ISO~21378:2019--compliant~\citep{iso21378} datasets covering all
industry sectors, each containing an entire fiscal year of general ledger
data.  The corpus totals approximately 364 million journal entries with
2.4 billion line items.  \Cref{tab:realworld_datasets} summarizes the
dataset classification.

\begin{table}[t]
\centering
\caption{Classification of 155 real-world general ledger datasets
  by size~\citep{ivertowski2024hardware}.}
\label{tab:realworld_datasets}
\small
\begin{tabular}{lccc}
\toprule
& \textbf{Small} & \textbf{Medium} & \textbf{Large} \\
& ($<$1M lines p.a.) & (1--10M lines p.a.) & ($>$10M lines p.a.) \\
\midrule
Number of datasets & 74 & 47 & 34 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Journal entry line item distribution.}
\Cref{tab:realworld_lineitems} shows the distribution of line items per
journal entry across the full corpus.  The distribution is extremely
heavy-tailed: 92.9\% of entries have 2--9 line items, while the maximum
observed is 47{,}059.  This motivates \system{}'s log-normal mixture
model for line-item count generation (\Cref{sec:stat_mixture}).

\begin{table}[t]
\centering
\caption{Distribution of journal entry line items across 155 real-world
  datasets~\citep{ivertowski2024hardware}.  The data exhibits extreme
  concentration in the 2--9 range.}
\label{tab:realworld_lineitems}
\small
\begin{tabular}{crrrr}
\toprule
\textbf{Bin} & \textbf{Total JEs} & \textbf{Coverage (\%)}
  & \textbf{Cum.\ Cov.\ (\%)} & \textbf{Line Items} \\
\midrule
1 & 338M & 92.90 & 92.90 & 2 to 9 \\
2 & 23M  & 6.33  & 99.22 & 10 to 99 \\
3 & 2.80M & 0.76 & 99.98 & 100 to 999 \\
4 & 63k  & 0.02  & 100.00 & 1{,}000 to 9{,}999 \\
5 & 196  & 0.00  & 100.00 & $\geq$ 10{,}000 \\
\bottomrule
\end{tabular}
\end{table}

\Cref{fig:je_line_items} provides a detailed breakdown within Bin~1.
The prominence of 2-line entries (60.7\%) and 4-line entries (16.6\%)
reflects the balanced nature of double-entry bookkeeping, where
equal numbers of debit and credit lines are the
norm~\citep{fellingham2018double}.  Overall, 88\% of journal entries have
an even number of line items, consistent with GAAP's requirement that
total debits equal total credits.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=14pt,
    xlabel={Line Items per Journal Entry},
    ylabel={Coverage (\%)},
    symbolic x coords={2,3,4,5,6,7,8,9},
    xtick=data,
    nodes near coords,
    nodes near coords align={vertical},
    every node near coord/.append style={font=\scriptsize},
    ymin=0, ymax=72,
    width=0.88\textwidth,
    height=6cm,
    ylabel style={at={(axis description cs:-0.06,.5)}},
    enlarge x limits=0.12,
]
\addplot[fill=blue!40, draw=blue!70] coordinates {
  (2,60.68) (3,5.77) (4,16.63) (5,3.06)
  (6,3.32) (7,1.13) (8,1.88) (9,0.42)
};
\end{axis}
\end{tikzpicture}
\caption{Detailed distribution of journal entry line items in the 2--9
  range (Bin~1, covering 92.9\% of all entries) from 155 real-world
  general ledger datasets~\citep{ivertowski2024hardware}.  The dominance
  of 2-line and 4-line entries reflects balanced double-entry bookkeeping.}
\label{fig:je_line_items}
\end{figure}

\paragraph{Chart of accounts complexity.}
The 155 datasets exhibit a wide range of chart-of-accounts complexity:
on average 557 financial accounts per entity (median: 367, min: 76,
max: 2{,}536).  \system{}'s three complexity tiers (small $\sim$100,
medium $\sim$400, large $\sim$2{,}500) were calibrated to span this
observed range, covering the 10th, 50th, and 99th percentiles of real-world
CoA sizes.

\paragraph{Graph construction complexity.}
The combinatorial complexity of constructing accounting networks
from journal entries is driven by the credit--debit matching problem.
For a journal entry with $n$ credit and $m$ debit line items, the number
of possible edge constructions is given by the Stirling number of the
second kind~\citep{ivertowski2024hardware}:
\begin{equation}
a = n!\;\frac{1}{m!}\sum_{j=0}^{m}(-1)^{m-j}\binom{m}{j}j^n
\label{eq:partition_complexity}
\end{equation}
In production data, \citet{ivertowski2024hardware} find that only 60.7\%
of entries admit trivial 1-to-1 mapping (2-line entries), while 39.3\%
require more sophisticated matching strategies.  \system{}'s graph export
module (\Cref{sec:gen_graph}) generates transaction networks that
preserve these structural characteristics, producing graphs with
realistic degree distributions, edge density, and community structure.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    bar width=10pt,
    xlabel={Matching Method},
    ylabel={Coverage (\%)},
    symbolic x coords={1:1 mapping, $n$:$n$ balanced, $n$:$m$ unequal,
      Higher agg., Decomposition},
    xtick=data,
    x tick label style={rotate=25, anchor=east, font=\small},
    nodes near coords,
    every node near coord/.append style={font=\scriptsize},
    ymin=0, ymax=72,
    width=0.88\textwidth,
    height=6cm,
    enlarge x limits=0.12,
]
\addplot[fill=green!40, draw=green!70] coordinates {
  ({1:1 mapping},60.68)
  ({$n$:$n$ balanced},22.22)
  ({$n$:$m$ unequal},9.58)
  ({Higher agg.},5.52)
  ({Decomposition},2.00)
};
\end{axis}
\end{tikzpicture}
\caption{Distribution of journal entry resolution methods required for
  accounting network construction, derived from 155 real-world
  datasets~\citep{ivertowski2024hardware}.  Methods A--E correspond to
  increasing combinatorial complexity.  \system{} generates entries
  with matching distributions across these categories.}
\label{fig:solving_methods}
\end{figure}

% ─────────────────────────────────────────────────────────────────────
\subsection{Statistical Fidelity}
\label{sec:exp_statistical}

\paragraph{Benford's Law Compliance.}
\Cref{tab:benford_results} reports first-digit MAD scores for journal-entry
amounts across configurations.  All scores fall within the ``close
conformity'' band ($\text{MAD} < 0.006$)~\citep{nigrini2012benfords}.

\begin{table}[t]
\centering
\caption{Benford first-digit MAD scores for generated journal-entry amounts.
  Threshold for close conformity: MAD~$< 0.006$.}
\label{tab:benford_results}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Preset} & \textbf{$n$ entries} & \textbf{MAD}
  & \textbf{$\chi^2$ $p$-value} \\
\midrule
Retail (Small)       & 52{,}000   & 0.0041 & 0.72 \\
Manufacturing (Med.) & 385{,}000  & 0.0023 & 0.91 \\
Financial (Large)    & 2{,}100{,}000 & 0.0011 & 0.98 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Distribution Fitting.}
Anderson-Darling tests for log-normal fit on transaction amounts yield
$p$-values $> 0.05$ across all configurations, confirming that the mixture
sampling preserves the target distributional form.  Percentile comparisons
(1st, 25th, 50th, 75th, 99th) show $< 2\%$ relative deviation from
configured parameters.

\paragraph{Correlation Preservation.}
For the Manufacturing preset with Gaussian copula ($\rho_{\text{amount,line\_items}} = 0.65$),
the empirical Pearson correlation is $\hat{\rho} = 0.648 \pm 0.003$ across
10 seeded runs, well within the tolerance threshold.

% ─────────────────────────────────────────────────────────────────────
\subsection{Structural Coherence}
\label{sec:exp_coherence}

\Cref{tab:coherence_results} summarizes the coherence validator results
for the Manufacturing (Medium) configuration.

\begin{table}[t]
\centering
\caption{Coherence validation results for Manufacturing (Medium) preset.}
\label{tab:coherence_results}
\small
\begin{tabular}{lcc}
\toprule
\textbf{Validator} & \textbf{Pass Rate} & \textbf{Violations} \\
\midrule
Balanced entries (debit $=$ credit)   & 100.0\% & 0 \\
Accounting equation ($A = L + E$)     & 100.0\% & 0 \\
Document chain referential integrity  & 100.0\% & 0 \\
Three-way match (within tolerance)    & 98.7\%  & 52 \\
Subledger $=$ GL control accounts     & 100.0\% & 0 \\
IC elimination nets to zero           & 100.0\% & 0 \\
Financial statement interconnection   & 100.0\% & 0 \\
Trial balance (debits $=$ credits)    & 100.0\% & 0 \\
\bottomrule
\end{tabular}
\end{table}

The 52 three-way match violations are \emph{intentional}---they correspond
to injected price and quantity variances that mirror real-world procurement
discrepancies.  All hard invariants (balanced entries, accounting equation,
subledger reconciliation) achieve 100\% compliance.

% ─────────────────────────────────────────────────────────────────────
\subsection{Downstream ML Utility}
\label{sec:exp_ml}

We evaluate whether \system{}-generated data supports effective fraud
detection model training.

\paragraph{Setup.}
We train three model families on the Financial Services (Large) dataset
with 5\% anomaly injection:
\begin{enumerate}
  \item \textbf{XGBoost:} Gradient-boosted trees on tabular features.
  \item \textbf{GCN:} Graph convolutional network on the transaction graph
    (PyTorch Geometric export).
  \item \textbf{Isolation Forest:} Unsupervised anomaly detection as a
    baseline.
\end{enumerate}

We use a temporal 70/15/15 train/validation/test split.  Evaluation
metrics: F1-score, AUPRC (area under precision-recall curve), and recall
at 5\% false positive rate (R@5\%FPR).

\paragraph{Results.}

\begin{table}[t]
\centering
\caption{Fraud detection performance on \system{}-generated data.}
\label{tab:ml_results}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Model} & \textbf{F1} & \textbf{AUPRC} & \textbf{R@5\%FPR} \\
\midrule
Isolation Forest   & 0.42 & 0.38 & 0.31 \\
XGBoost            & 0.81 & 0.79 & 0.73 \\
GCN                & 0.84 & 0.82 & 0.78 \\
\bottomrule
\end{tabular}
\end{table}

The GCN model achieves the highest performance, benefiting from the
relational structure preserved in \system{}'s graph exports.  The gap
between supervised (XGBoost, GCN) and unsupervised (Isolation Forest)
methods confirms that the generated labels provide meaningful signal.

\paragraph{Difficulty Stratification.}
When we stratify test-set performance by anomaly difficulty score
(\Cref{sec:anom_scoring}), we observe the expected gradient: easy
anomalies (difficulty $< 0.3$) are detected at $>$95\% recall, while
hard anomalies (difficulty $> 0.7$) yield $\sim$45\% recall, validating
that the difficulty scoring produces a meaningful proxy for detectability.

% ─────────────────────────────────────────────────────────────────────
\subsection{System Performance}
\label{sec:exp_performance}

\paragraph{Throughput.}

\begin{table}[t]
\centering
\caption{Generation throughput (single-threaded) by record type.}
\label{tab:throughput}
\small
\begin{tabular}{lr}
\toprule
\textbf{Record Type} & \textbf{Throughput (records/sec)} \\
\midrule
Journal entries           & 112{,}000 \\
Journal line items        & 340{,}000 \\
Master data (combined)    & 58{,}000 \\
Document-flow records     & 85{,}000 \\
Subledger records         & 95{,}000 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Distribution Sampling.}
Micro-benchmarks for individual samplers:

\begin{center}
\small
\begin{tabular}{lr}
\toprule
\textbf{Sampler} & \textbf{Latency} \\
\midrule
BenfordSampler          & $<1\,\mu$s / sample \\
Copula (Gaussian, 3-D)  & 2--5\,$\mu$s / sample \\
LogNormalMixture (3-comp.) & $<500$\,ns / sample \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Output Sink Performance.}

\begin{center}
\small
\begin{tabular}{lr}
\toprule
\textbf{Format} & \textbf{Throughput (records/sec)} \\
\midrule
CSV (streaming)   & 210{,}000 \\
JSON (streaming)  & 105{,}000 \\
Parquet (columnar) & 520{,}000 \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Scaling.}
Throughput scales near-linearly with core count up to 8 cores (due to the
per-generator seed isolation described in \Cref{sec:arch_determinism}),
after which contention on shared data structures (balance tracker,
document-chain manager) introduces sublinear scaling.  At 16 cores, the
Manufacturing (Medium) configuration generates a full 24-month dataset
(385K journal entries, 1.1M line items, 500K+ document-flow records)
in under 30 seconds wall-clock time.

\paragraph{Memory Efficiency.}
The streaming architecture keeps peak resident memory below 500\,MB even
for the Financial Services (Large) configuration.  The graceful degradation
system (\Cref{sec:arch_resources}) was not triggered during any benchmark
run on the 64\,GB test machine.

% ─────────────────────────────────────────────────────────────────────
\subsection{Privacy-Utility Evaluation}
\label{sec:exp_privacy}

We evaluate the fingerprint module by extracting a fingerprint from the
Manufacturing (Medium) dataset at each privacy level, synthesizing a new
dataset from each fingerprint, and comparing statistical properties.

\begin{table}[t]
\centering
\caption{Fingerprint privacy-utility tradeoff.  KL divergence between
  original and fingerprint-synthesized amount distributions.}
\label{tab:privacy_utility}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Privacy Level} & $\varepsilon$ & \textbf{KL Divergence}
  & \textbf{Benford MAD} \\
\midrule
Minimal  & 5.0 & 0.003 & 0.0026 \\
Standard & 1.0 & 0.008 & 0.0039 \\
High     & 0.5 & 0.019 & 0.0052 \\
Maximum  & 0.1 & 0.071 & 0.0098 \\
\bottomrule
\end{tabular}
\end{table}

Even at the ``High'' privacy level ($\varepsilon = 0.5$), the
KL divergence remains below 0.02 and Benford MAD stays within close
conformity, confirming the practicality of the fingerprint approach
for enterprise-scale data.
