% sections/generation_pipeline.tex — Generation Pipeline
\section{Generation Pipeline}
\label{sec:generation_pipeline}

\system{}'s generation pipeline transforms a declarative YAML configuration
into a complete, internally consistent enterprise dataset.  This section
details the pipeline stages and the business-process generators that produce
coherent multi-table output.

% ─────────────────────────────────────────────────────────────────────
\subsection{Orchestration Workflow}
\label{sec:gen_orchestration}

The \texttt{GenerationOrchestrator} executes the following stages in order:

\begin{enumerate}
  \item \textbf{Configuration loading and validation.}  Parse YAML, apply
    industry preset defaults, validate invariants (balanced weights,
    ascending thresholds, valid date ranges).

  \item \textbf{Chart of Accounts (CoA) initialization.}  Generate a GL
    account structure at the configured complexity level (small: $\sim$100,
    medium: $\sim$400, large: $\sim$2{,}500 accounts) with proper account
    hierarchies, types (Asset, Liability, Equity, Revenue, Expense), and
    control accounts (AR~$=$ 1100, AP~$=$ 2000).

  \item \textbf{Master data generation.}  Produce vendors, customers,
    materials, fixed assets, employees, and cost centers with
    cross-referenced relationships (e.g., vendor bank details, customer
    credit limits, material-vendor linkages).

  \item \textbf{Opening balance generation.}  Establish period-start balances
    satisfying the accounting equation $A = L + E$.  The balance tracker
    maintains running account-level balances throughout subsequent stages.

  \item \textbf{Document-flow generation.}  Execute P2P, O2C, S2C, HR,
    and manufacturing process generators that produce linked document chains
    and corresponding journal entries.

  \item \textbf{Period-close processing.}  Run accruals, depreciation,
    FX translation, intercompany elimination, and year-end closing entries.
    Generate trial balances and financial statements.

  \item \textbf{Anomaly injection.}  Apply configured fraud, error, process,
    statistical, and relational anomalies with ground-truth labels.

  \item \textbf{Data quality variation.}  Optionally inject missing values,
    format variations, typos, and encoding issues.

  \item \textbf{Evaluation.}  Run coherence validators, statistical tests,
    and ML-readiness checks.  The auto-tuner may propose configuration
    patches.

  \item \textbf{Output.}  Stream results to configured sinks (CSV, JSON,
    Parquet) with optional compression.
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────
\subsection{Journal Entry Generation}
\label{sec:gen_journal}

Journal entries are the atomic unit of accounting data.  Each entry consists
of a header (ID, date, posting date, company code, user, description) and
one or more line items (account, amount, debit/credit indicator, cost center,
profit center).

\begin{definition}[Balanced Entry]
A journal entry $j$ with line items $\{l_1, \ldots, l_n\}$ is
\emph{balanced} iff:
\begin{equation}
\sum_{i : l_i.\text{debit}} l_i.\text{amount}
  = \sum_{i : \neg l_i.\text{debit}} l_i.\text{amount}.
\label{eq:balanced}
\end{equation}
\end{definition}

\system{} enforces \Cref{eq:balanced} at construction time: the
\texttt{JournalEntry} constructor returns an error if the debit-credit
difference exceeds a configurable tolerance (default: \$0.01).  This
invariant is maintained across all generators---no unbalanced entry can
enter the output pipeline.

Line-item amounts are drawn from the configured mixture distribution
(\Cref{eq:lognormal_mixture}).  The number of line items per entry follows
a correlated draw from the copula model (\Cref{sec:stat_copula}), ensuring
that higher-amount entries tend to have more line items and require higher
approval levels.

% ─────────────────────────────────────────────────────────────────────
\subsection{Document Flow Generation}
\label{sec:gen_docflow}

Enterprise transactions span multiple documents with strict sequencing
and referential integrity.

\paragraph{Procure-to-Pay (P2P).}
The P2P generator produces the chain:
\[
\text{Purchase Order} \to \text{Goods Receipt} \to \text{Vendor Invoice}
\to \text{Payment}
\]
Each stage creates the corresponding document record, updates the
document-chain manager with forward/backward references, and posts
journal entries to the appropriate GL accounts (inventory, GR/IR clearing,
AP, cash).

\paragraph{Three-Way Matching.}
The three-way matcher validates consistency across PO, GR, and invoice:

\begin{itemize}
  \item \textbf{Quantity tolerance:} $|q_{\text{PO}} - q_{\text{GR}}| \leq
    \epsilon_q \cdot q_{\text{PO}}$ (default $\epsilon_q = 2\%$).
  \item \textbf{Price tolerance:} $|p_{\text{PO}} - p_{\text{inv}}| \leq
    \epsilon_p \cdot p_{\text{PO}}$ (default $\epsilon_p = 5\%$).
  \item \textbf{Over-delivery:} Permitted up to $\delta_{\max} = 10\%$
    above PO quantity.
\end{itemize}

Variance types (quantity PO-GR, quantity GR-invoice, price PO-invoice,
total amount, missing line, extra line) are tracked and available as
evaluation targets.

\paragraph{Order-to-Cash (O2C).}
The O2C generator produces:
\[
\text{Sales Order} \to \text{Delivery} \to \text{Customer Invoice}
\to \text{Customer Receipt}
\]
with corresponding revenue recognition entries, AR postings, and cash
receipts.  Partial deliveries and split invoicing are supported.

\paragraph{Source-to-Contract (S2C).}
A full sourcing lifecycle: sourcing projects, supplier qualifications,
RFx events, supplier bids, bid evaluations, procurement contracts,
catalog items, and supplier scorecards.

\paragraph{HR/Payroll.}
Payroll runs with line items (base pay, overtime, deductions, taxes),
time entries, and expense reports with configurable approval workflows.
A benefit enrollment generator tracks employee plan selections
(health, dental, vision, retirement) with employer and employee
contribution amounts, enrollment windows, and coverage periods.

\paragraph{Manufacturing.}
Production orders, routing operations, quality inspections with
characteristics, and cycle counts.  A bill-of-materials (BOM) component
generator produces multi-level product structures with phantom assembly
support and quantity-per calculations.  Inventory movement tracking
covers receipt, issue, transfer, return, scrap, and adjustment movement
types, each posting corresponding material-ledger journal entries.

% ─────────────────────────────────────────────────────────────────────
\subsection{Subledger and Period-Close Processing}
\label{sec:gen_subledger}

\paragraph{Subledger generation.}
Four subledger generators maintain detailed subsidiary records:
\begin{itemize}
  \item \textbf{AR subledger:} Customer invoices, receipts, aging buckets,
    credit memos, adjustments.
  \item \textbf{AP subledger:} Vendor invoices, payments, aging, credit
    notes, prepayments.
  \item \textbf{FA subledger:} Asset capitalization, depreciation schedules
    (straight-line, declining balance), disposals, impairment.
  \item \textbf{Inventory subledger:} Receipts, issues, valuation
    (FIFO/LIFO/weighted average), write-downs.
\end{itemize}

Each subledger reconciles to its control account in the GL (e.g., AR
subledger total $=$ GL account 1100).

\paragraph{Period-close engine.}
The close engine executes:
(i)~accrual entries for unbilled revenue/expenses,
(ii)~depreciation postings from the FA subledger,
(iii)~FX translation using period-end spot rates with CTA postings,
(iv)~intercompany matching and elimination,
(v)~year-end income-to-equity reclassification,
and (vi)~financial statement generation (balance sheet, income statement,
cash flow, changes in equity).

% ─────────────────────────────────────────────────────────────────────
\subsection{Intercompany Processing}
\label{sec:gen_intercompany}

Multi-entity datasets require intercompany (IC) transaction matching and
elimination for consolidated reporting.  The IC generator creates bilateral
transaction pairs between related entities.  The matching engine identifies
matched pairs within configurable tolerances.  The elimination generator
produces consolidation entries that net IC receivables/payables, IC
revenue/expense, and unrealized profit in inventory.

% ─────────────────────────────────────────────────────────────────────
\subsection{Graph Export}
\label{sec:gen_graph}

Double-entry bookkeeping has a natural directed-graph
representation~\citep{ijiri1965generalized}: each distinct GL account
forms a node, credit entries produce outgoing edges, and debit entries
produce incoming edges.  Following~\citet{ivertowski2024hardware}, we
adopt a temporal formulation:
\begin{equation}
G(t_0, t_1) = A(t_0, t_1) \times E(t_0, t_1)
\label{eq:temporal_graph}
\end{equation}
where $A(t_0,t_1)$ is the set of account nodes with monetary state over
$[t_0, t_1]$, and each edge $E_i(t_0,t_1) = \sum_{n=1}^{|\Omega_{A_i}|}
\omega_n$ aggregates all transactions $\omega_n$ affecting node $A_i$ in
that period.  This temporal definition enables system-theoretic analyses
such as cross-correlation and auto-correlation on account flows.

\system{} constructs heterogeneous graphs from the generated data and
exports them in formats suitable for graph neural network (GNN) research:

\begin{itemize}
  \item \textbf{Transaction graph:} Nodes are accounts/entities; edges are
    transactions with amount, date, and process-type features.
  \item \textbf{Approval network:} Nodes are users; edges are approval
    relationships with threshold attributes.  Segregation-of-duties
    conflicts are detectable as structural motifs.
  \item \textbf{Entity relationship graph:} Nodes are legal entities,
    vendors, customers; edges represent ownership, control, and supply-chain
    relationships.
\end{itemize}

Export formats: PyTorch Geometric (\texttt{.pt} tensors), Neo4j (CSV +
Cypher scripts), and DGL (heterograph with typed nodes/edges).

As of v0.10, the graph module covers over 50 entity types across ten
process families (Tax, Treasury, ESG, Project Accounting, Source-to-Contract,
Hire-to-Retire, Manufacturing, Governance, and core P2P/O2C), connected by
28~relationship edge types with explicit cardinality constraints (e.g.,
one-to-many for invoice-to-line-item, many-to-many for vendor-to-material).
A typed \texttt{ToNodeProperties} trait maps domain models to seven
property-value types (\texttt{String}, \texttt{Decimal}, \texttt{Integer},
\texttt{Boolean}, \texttt{Date}, \texttt{DateTime}, \texttt{List}),
ensuring lossless round-tripping through graph serialization.  A
hypergraph builder extends the model to multi-entity relationships---for
example, a three-way match hyperedge linking a purchase order, goods
receipt, and vendor invoice simultaneously.

% ─────────────────────────────────────────────────────────────────────
\subsection{Process Mining (OCEL 2.0)}
\label{sec:gen_ocel}

The OCPM module generates Object-Centric Event Logs conforming to the
OCEL~2.0 standard~\citep{ghahfarokhi2021ocel}.  Events are linked to
multiple typed objects (Order, Invoice, Shipment, Payment) via many-to-many
relationships.  Process variants (standard P2P, P2P with three-way match
variance, full O2C, O2C with returns) are generated with empirically
calibrated frequency distributions.  The resulting logs are directly
importable into process mining tools such as PM4Py and Celonis.

Version 0.11 enriches the OCEL module with lifecycle state machines for
\texttt{PurchaseOrder}, \texttt{SalesOrder}, and \texttt{VendorInvoice}
objects.  Each state machine defines probabilistic transitions (e.g.,
\emph{Created} $\to$ \emph{Approved} with $p=0.92$, \emph{Created} $\to$
\emph{Rejected} with $p=0.08$), producing realistic variant distributions
without manual scripting.  Multi-object correlation events---three-way
match, payment allocation, and bank reconciliation---link objects across
process boundaries in a single event, reflecting the cross-functional
nature of enterprise operations.  Resource pool workload modeling assigns
events to users via configurable strategies (round-robin, least-busy,
skill-based), and enriched events carry \texttt{from\_state},
\texttt{to\_state}, \texttt{resource\_workload}, and
\texttt{correlation\_id} attributes for downstream conformance checking
and resource-perspective analysis.
