% sections/privacy.tex — Privacy-Preserving Fingerprint Module
\section{Privacy-Preserving Fingerprint Module}
\label{sec:privacy}

A key challenge in synthetic data generation is preserving the statistical
characteristics of real production data without exposing individual records.
\system{}'s fingerprint module addresses this by extracting
\emph{distributional summaries} under formal privacy guarantees, producing
a compact archive from which a fully configured generator can be instantiated.

% ─────────────────────────────────────────────────────────────────────
\subsection{Privacy Mechanisms}
\label{sec:priv_mechanisms}

The module combines two complementary privacy techniques:

\paragraph{Differential Privacy ($\varepsilon$-DP).}
Each extracted statistic is protected by calibrated Laplace noise:

\begin{equation}
\tilde{s} = s + \text{Lap}\!\left(\frac{\Delta s}{\varepsilon}\right)
\label{eq:dp_laplace}
\end{equation}

where $s$ is the true statistic, $\Delta s$ is its sensitivity (maximum
change from adding or removing one record), and $\varepsilon$ is the privacy
budget.  The module tracks cumulative $\varepsilon$ expenditure across all
queries via a privacy accountant, ensuring that the total privacy loss remains
within the configured budget.

Four privacy levels provide pre-configured $\varepsilon$ values:

\begin{center}
\begin{tabular}{llccc}
\toprule
\textbf{Level} & \textbf{$\varepsilon$} & \textbf{$k$-anon}
  & \textbf{Privacy} & \textbf{Utility} \\
\midrule
Minimal  & 5.0 & 3  & Low    & High \\
Standard & 1.0 & 5  & Medium & Medium \\
High     & 0.5 & 10 & High   & Reduced \\
Maximum  & 0.1 & 20 & Maximum & Low \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{$k$-Anonymity.}
Categorical distributions with fewer than $k$ occurrences in any group
are suppressed (generalized to a parent category or marked as ``Other'').
This prevents re-identification of rare attribute combinations.

% ─────────────────────────────────────────────────────────────────────
\subsection{Fingerprint Extraction}
\label{sec:priv_extraction}

The \texttt{FingerprintExtractor} processes input data (CSV, Parquet) in a
streaming fashion to minimize memory usage.  Seven extraction components run
sequentially:

\begin{enumerate}
  \item \textbf{Schema extraction:} Table names, column types, nullable
    flags, foreign-key relationships.
  \item \textbf{Statistics extraction:} Per-column distribution parameters
    (mean, variance, percentiles 1/5/25/50/75/95/99), Benford digit
    frequencies, cardinality estimates.
  \item \textbf{Correlation extraction:} Pairwise correlation matrices,
    copula family selection (via AIC/BIC), fitted copula parameters.
  \item \textbf{Integrity extraction:} Foreign-key cardinality distributions,
    referential integrity rates, orphan record counts.
  \item \textbf{Rule extraction:} Balance constraints (debit $=$ credit),
    approval thresholds, business rules.
  \item \textbf{Anomaly extraction:} Anomaly rates by category, type
    distribution, temporal concentration.
  \item \textbf{Privacy audit:} Records all privacy decisions, noise
    additions, suppression counts, and total $\varepsilon$ spent.
\end{enumerate}

Each numeric statistic is noised via \Cref{eq:dp_laplace} before being
written to the fingerprint.

% ─────────────────────────────────────────────────────────────────────
\subsection{Fingerprint File Format (.dsf)}
\label{sec:priv_format}

The fingerprint is stored as a ZIP archive (\texttt{.dsf}) containing
eight YAML/JSON components:

\begin{center}
\small
\begin{tabular}{ll}
\toprule
\textbf{File} & \textbf{Contents} \\
\midrule
\texttt{manifest.json}      & Version, checksums, privacy config, signature \\
\texttt{schema.yaml}        & Tables, columns, types, FK relationships \\
\texttt{statistics.yaml}    & Distributions, percentiles, Benford frequencies \\
\texttt{correlations.yaml}  & Correlation matrices, copula parameters \\
\texttt{integrity.yaml}     & FK cardinality, constraints \\
\texttt{rules.yaml}         & Balance constraints, thresholds \\
\texttt{anomalies.yaml}     & Anomaly rates, type distribution \\
\texttt{privacy\_audit.json} & Privacy decisions, $\varepsilon$ spent \\
\bottomrule
\end{tabular}
\end{center}

The \texttt{manifest.json} includes SHA-256 checksums for each component,
enabling tamper detection.  The \texttt{privacy\_audit.json} provides a
complete audit trail of all privacy-related decisions made during extraction.

% ─────────────────────────────────────────────────────────────────────
\subsection{Configuration Synthesis}
\label{sec:priv_synthesis}

The \texttt{ConfigSynthesizer} reads a \texttt{.dsf} fingerprint and
generates a complete \system{} YAML configuration.  Key synthesis steps:

\begin{enumerate}
  \item \textbf{Scale estimation:} Target row counts are derived from the
    original counts multiplied by a configurable factor (default $2.0\times$).
  \item \textbf{Distribution mapping:} Noised statistics are mapped to the
    nearest supported distribution family (log-normal mixture, Gaussian
    mixture, Pareto, etc.) via method-of-moments or maximum likelihood.
  \item \textbf{Copula reconstruction:} Correlation matrices and copula
    parameters are used to configure the dependency model.
  \item \textbf{Anomaly calibration:} Observed anomaly rates are mapped to
    injection parameters.
  \item \textbf{Seed assignment:} A new deterministic seed is generated,
    ensuring the synthesized data is independent of the original.
\end{enumerate}

This two-phase architecture (extract $\to$ synthesize) cleanly separates the
\emph{privacy boundary} (extraction, where DP is applied) from the
\emph{generation process} (synthesis, which is purely algorithmic), enabling
independent auditing and regulatory review.

% ─────────────────────────────────────────────────────────────────────
\subsection{Privacy-Utility Tradeoff}
\label{sec:priv_tradeoff}

\begin{proposition}[Utility bound]
\label{prop:utility}
Let $\tilde{\mu}$ be the noised mean of a column with $n$ records and
true mean $\mu$.  The sensitivity is $\Delta\mu = R/n$ where $R$ is the
column range.  Then:
\[
\E[|\tilde{\mu} - \mu|] = \frac{R}{n\varepsilon}
\]
For $n = 10{,}000$ records, $R = 10^6$ (typical transaction amount range),
and $\varepsilon = 1.0$: $\E[|\tilde{\mu} - \mu|] = 100$, which is $<0.01\%$
of the range---negligible for distribution fitting purposes.
\end{proposition}

This result demonstrates that for enterprise-scale datasets (typically
$n > 10{,}000$ records), even standard privacy levels ($\varepsilon = 1.0$)
introduce negligible utility loss for aggregate statistics, making the
fingerprint approach practical for real-world deployments.

% ─────────────────────────────────────────────────────────────────────
\subsection{Workflow}
\label{sec:priv_workflow}

The end-to-end privacy-preserving workflow is:

\begin{enumerate}
  \item \textbf{Extract:}
    \texttt{datasynth-data fingerprint extract --input data.csv --output fp.dsf --privacy-level standard}
  \item \textbf{Validate:}
    \texttt{datasynth-data fingerprint validate fp.dsf} (verify integrity,
    inspect privacy audit)
  \item \textbf{Generate:}
    \texttt{datasynth-data generate --fingerprint fp.dsf --output ./synthetic/}
  \item \textbf{Evaluate:}
    \texttt{datasynth-data fingerprint evaluate --fingerprint fp.dsf --synthetic ./synthetic/}
\end{enumerate}

The fingerprint file can be safely shared across organizational boundaries,
as it contains only noised aggregate statistics---never individual records.
