% sections/related_work.tex — Related Work
\section{Related Work}
\label{sec:related_work}

We position \system{} at the intersection of five research streams:
enterprise knowledge systems and audit analytics, synthetic tabular data
generation, accounting-domain data synthesis, anomaly injection for fraud
detection, and privacy-preserving data sharing.

% ─────────────────────────────────────────────────────────────────────
\subsection{Enterprise Knowledge Systems for Audit}
\label{sec:rw_knowledge}

The construction of knowledge graphs from enterprise data has gained
increasing attention in audit analytics.  Existing approaches follow
the inverse recovery paradigm described in \Cref{sec:gt_inverse}:
they extract entities and relationships from operational data to build
audit-relevant knowledge structures.

\citet{guo2022picture} demonstrate that accounting graph
topology---degree distributions, clustering coefficients, and community
structure---can discriminate between normal and fraudulent bookkeeping
patterns, validating the information content of structural knowledge
(Layer~1 in our framework).  \citet{boersma2024complex} applies complex
network analysis to audit data, modeling interfirm relationships and
transaction patterns as multi-layer graphs.
\citet{liang2021pattern} develop pattern recognition and anomaly detection
methods directly on bookkeeping graph representations.

These approaches generally rely on available operational data as
their starting point and do not separately establish that the input
faithfully represents the underlying business reality.  As discussed
in \Cref{sec:gt_errors}, when systematic errors are present---precisely
the situation where audit analytics is most critical---this reliance
can limit the reliability of the resulting knowledge structures.
\system{} complements these approaches by providing reference knowledge
graphs with known ground truth, enabling rigorous evaluation of
knowledge construction methods.

% ─────────────────────────────────────────────────────────────────────
\subsection{General-Purpose Tabular Data Generation}
\label{sec:rw_tabular}

Deep generative models have become the dominant paradigm for synthetic
tabular data.  CTGAN~\citep{xu2019modeling} introduced mode-specific
normalization and conditional training for mixed-type columns.
TVAE~\citep{xu2019modeling} applies variational autoencoders with the
same preprocessing.  Subsequent work has refined these approaches:
CopulaGAN~\citep{patki2016synthetic} combines copula transformations
with GAN training, REaLTabFormer~\citep{solatorio2023realtabformer}
leverages autoregressive transformers for relational tables, and
TabDDPM~\citep{kotelnikov2023tabddpm} adapts diffusion models to
tabular settings.

While these methods excel at preserving marginal distributions and
pairwise correlations, the enterprise accounting domain imposes additional requirements
beyond what general-purpose tabular generators address: cross-table
referential integrity, temporal ordering constraints, and
domain-specific algebraic invariants (e.g., balanced entries).
Furthermore, audit analytics requires a normative knowledge
layer---accounting standards, controls, and regulatory compliance---that
is beyond the scope of general-purpose generators.

% ─────────────────────────────────────────────────────────────────────
\subsection{Accounting-Domain Data Synthesis}
\label{sec:rw_accounting}

Domain-specific synthetic data for accounting has received comparatively
little attention.  \citet{schreyer2022federated} propose a federated
learning framework for training anomaly detection models on journal
entry data across audit engagements, addressing data-sharing constraints
but not the generation of synthetic reference data.
BankSim~\citep{lopez2014banksim} focuses narrowly on payment
transactions for fraud research.

These systems were designed to address specific aspects of the
synthetic data challenge and serve their intended purposes
effectively.  However, they do not produce a comprehensive reference
knowledge graph with provenance across all three layers, which is what
evaluating end-to-end knowledge construction pipelines requires.

% ─────────────────────────────────────────────────────────────────────
\subsection{Accounting Network Construction}
\label{sec:rw_networks}

The representation of double-entry bookkeeping as directed graphs dates
back to~\citet{ijiri1965generalized}, who established that journal entries
have a natural graph interpretation: accounts are nodes, and debit/credit
flows form directed edges.  \citet{fellingham2018double} provides a modern
algebraic treatment of double-entry systems.
\citet{arya2000inferring} study the problem of inferring individual
transactions from aggregated financial statements, highlighting the
combinatorial complexity of reconstructing edge-level flows from
node-level balances---a result closely related to our
\Cref{thm:combinatorial}.

\citet{ivertowski2024hardware} present a hardware-accelerated approach
for constructing accounting networks from general ledger data, analyzing
155 real-world ISO~21378:2019--compliant datasets comprising
approximately 364 million journal entries with 2.4 billion line items.
Their analysis reveals the distributional characteristics and
combinatorial complexity that directly inform \system{}'s generation
model and motivate the forward-generation approach.

While these approaches focus on \emph{analyzing} real accounting data as
graphs, none address the complementary problem of \emph{generating}
synthetic accounting data that produces reference knowledge graphs with
known ground truth.  \system{} bridges this gap.

% ─────────────────────────────────────────────────────────────────────
\subsection{Anomaly Injection and Fraud Simulation}
\label{sec:rw_anomaly}

Fraud detection benchmarks typically rely on post-hoc label injection
or manual scenario creation.  The IEEE-CIS fraud dataset~\citep{ieee_cis2019}
provides real transaction labels but lacks accounting context.
\citet{pourhabibi2020fraud} survey fraud detection methods and note the
scarcity of labeled multi-dimensional accounting fraud data.
\citet{dbouk2022anomaly} inject statistical anomalies into time series
but do not model multi-stage fraud schemes that evolve over time.

\system{} introduces a \emph{scheme-based} injection model where fraud
evolves through state-machine stages (e.g., vendor kickback schemes that
escalate over months), producing temporally correlated, multi-transaction
anomaly patterns with ground-truth labels spanning all five knowledge
dimensions (\Cref{def:dimensions}).

% ─────────────────────────────────────────────────────────────────────
\subsection{Privacy-Preserving Data Sharing}
\label{sec:rw_privacy}

Differential privacy~\citep{dwork2006calibrating} provides formal
guarantees for statistical queries.  DPGAN~\citep{xie2018differentially}
and PATE-GAN~\citep{jordon2019pate} train generative models under DP
constraints but are computationally expensive and may degrade statistical
fidelity.  \system{}'s fingerprint module takes a different approach:
it extracts distributional summaries under $\varepsilon$-DP,
producing a compact archive from which a fully configured generator
can be instantiated, cleanly separating the privacy boundary from the
generation process.

% ─────────────────────────────────────────────────────────────────────
\subsection{Process Mining and Event Logs}
\label{sec:rw_process_mining}

Object-centric event logs (OCEL~2.0)~\citep{ghahfarokhi2021ocel}
generalize traditional single-case-ID event logs by associating events
with multiple typed objects.  While tools such as
PM4Py~\citep{berti2019process} support OCEL~2.0 import/export,
generating realistic synthetic OCEL~2.0 logs with consistent underlying
business data has not been addressed.  \system{} produces OCEL~2.0
event logs that are traceable to the same journal entries, document
flows, and master-data records that constitute the rest of the reference
knowledge graph.

% ─────────────────────────────────────────────────────────────────────
\subsection{Summary of Gaps}
\label{sec:rw_summary}

\Cref{tab:related_comparison} summarizes feature coverage.  No existing
system produces a complete reference knowledge graph spanning all three
knowledge layers with full provenance.

\begin{table}[t]
\centering
\caption{Feature comparison with related systems. Columns:
  \textbf{KG}~=~reference knowledge graph,
  \textbf{3L}~=~three knowledge layers,
  \textbf{Bal}~=~balanced entries,
  \textbf{Ben}~=~Benford compliance,
  \textbf{Doc}~=~document-flow chains,
  \textbf{Sub}~=~subledger coherence,
  \textbf{Std}~=~accounting/audit standards,
  \textbf{Anom}~=~anomaly injection with labels,
  \textbf{Graph}~=~graph export for GNNs.}
\label{tab:related_comparison}
\small
\begin{tabular}{l ccccccccc}
\toprule
\textbf{System} & \textbf{KG} & \textbf{3L} & \textbf{Bal} & \textbf{Ben}
  & \textbf{Doc} & \textbf{Sub} & \textbf{Std}
  & \textbf{Anom} & \textbf{Graph} \\
\midrule
CTGAN / TVAE          & -- & -- & -- & -- & -- & -- & -- & -- & -- \\
SDV / CopulaGAN       & -- & -- & -- & -- & \checkmark$^*$ & -- & -- & -- & -- \\
REaLTabFormer         & -- & -- & -- & -- & \checkmark & -- & -- & -- & -- \\
BankSim               & -- & -- & -- & -- & -- & -- & -- & \checkmark & -- \\
Schreyer et al.       & -- & -- & -- & -- & -- & -- & -- & -- & -- \\
Guo et al.            & -- & \checkmark$^*$ & -- & -- & -- & -- & -- & -- & \checkmark$^*$ \\
\midrule
\textbf{\system{}}    & \checkmark & \checkmark & \checkmark & \checkmark
  & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
\bottomrule
\multicolumn{10}{l}{\footnotesize $^*$Partial support.}
\end{tabular}
\end{table}
